A spiral galaxy's mass distribution uncovered through lensing and dynamics

### 2016-06-03 LensingPaper_draft_v4.01.tex
Finished editing of section on lensing theory. Started editing section on JAM theory.

%--------------------------------------------
\Wilma{[Referee says: "The manuscript in its current form is also slightly longer than necessary at certain
points: for instance, in section 3.2.1, it is not really necessary to present all
the lensing equations (e.g. the time delay, equation 7)"]}
A gravitational lens is a mass distribution whose gravitational potential $\Phi$ acts as a lens for light coming from a source positioned somewhere on a plane behind the lens. We summarize the basic formalism (see also \citealt{1992grle.book.....S,1996astro.ph..6001N,2006glsw.conf....1S,2006glsw.conf...91K,2010ARA&A..48...87T}). The angular diameter distance from the observer to the lens is $D_d$, to the source plane $D_s$, and the distance between the lens and source plane is $D_{ds}$. The deflection potential of the lens is its potential projected along the line of sight $z'$ and rescaled to
\begin{equation}
\psi(\vect{\theta}) := \frac{D_{ds}}{D_d D_s} \frac{2}{c^2} \int \Phi(\vect{r}=D_d \vect{\theta},z') {\ \mathrm d} z', \label{eq:psidef}
\end{equation}
where $\vect{\theta}\equiv(x',y')$ is a 2-dimensional vector on the plane of the sky. The light from the source at $\vect{\beta} \equiv (x_s',y_s')$ is deflected according to the lens equation
\begin{equation}
\vect{\beta} = \vect{\theta}_i - \left.\vect{\nabla}_\theta \psi(\vect{\theta})\right|_{\vect{\theta}_i} \label{eq:lenseqpot}
\end{equation}
into an image $\vect{\theta}_i = (x_i',y_i')$. The gradient of the deflection potential, $\vect{\nabla}_\theta \psi(\vect{\theta})$, is equal to the angle by which the light is deflected multiplied by $D_{ds}/D_{s}$.

@@

The inverse magnification tensor
\begin{equation}
\mathscr{M}^{-1} \equiv \frac{\partial \vect{\beta}}{\partial \vect{\theta}} \overset{(\ref{eq:lenseqpot})}{=} \left(\delta_{ij} - \frac{\partial^2 \psi}{\partial \theta_i \partial \theta_j} \right)\label{eq:magnificationtensor}
\end{equation}
describes how the source position changes with image position. It also describes the distortion of the image shape for an extended source and its magnification $\mu$ due to lensing according to
$$\mu \equiv \frac{\text{image area}}{\text{source area}} = \det \mathscr{M}.$$
Lines in the image plane for which the magnification becomes infinite, i.e., $\det \mathscr{M}^{-1} = 0$, are called critical curves. The corresponding lines in the source plane are called caustics. The position of the source with respect to the caustic detemines the number of images and their configuration and shape with respect to each other.

The Einstein mass $M_\text{ein}$ and Einstein radius $R_\text{ein}$ are defined via the relation
\begin{equation*}
M_\text{ein} \equiv M(<R_\text{ein}) \overset{!}{=} \pi \Sigma_\text{crit} R_\text{ein}^2,
\end{equation*}
where $\Sigma_\text{crit} \equiv \frac{c^2}{4\pi G} \frac{D_s}{D_d D_{ds}}$ is the critical density and $M(<R_\text{ein})$ is the mass projected along the line-of-sight within $R'=R_\text{ein}$. $M_\text{ein}$ is similar to the projected mass within the critical curve $M_\text{crit}$.
%--------------------------------------------

@@

%--------------------------------------------
\Wilma{[Referee says: "The manuscript in its current form is also slightly longer than necessary at certain
points: [...] section
3.3.1 contains a lot of material that is implicit in Cappellari's JAM code, and
therefore does not need to be explained equation by equation here."]}
JAM modelling assumes galaxies to be (i) collisionless (i.e., the collisionless Boltzmann equation for the distribution function $f(\vect{x},\vect{v},t)$ has to be satisfied, $\frac{\diff f(\vect{x},\vect{v},t)}{\diff t} = 0$), (ii) in a steady state ($\frac{\partial}{\partial t} = 0$), (iii) axisymmetric (best described in cylindrical coordinates $(R,z,\phi)$ and $\frac{\partial}{\partial \phi} = 0$). From this follow the axisymmetric Jeans equations as the vector-valued first moment of the Boltzmann equation, i.e.,
\begin{equation*}
\int \vect{v} \frac{\diff f}{\diff t} \Diff3 v = 0.
\end{equation*}
To be able to solve the Jeans equations, additional assumptions about the velocity ellipsoid tensor $\langle v_i v_j\rangle$ have to be made. We follow \citet{Cap08} and assume firstly that the galaxy's velocity ellipsoid is aligned with the cylindrical coordinate system, i.e., $\langle v_R v_z\rangle = \langle v_z v_\phi \rangle = \langle v_\phi v_R \rangle = 0$. Secondly, we assume a constant ratio between the radial and vertical second velocity moments, 
\begin{equation}
\beta_z \equiv 1 - \langle v_z^2 \rangle / \langle v_R^2\rangle, \label{eq:bz}
\end{equation}
where $\beta_z$ is the velocity anisotropy parameter. This reduces the Jeans equations to two equations for $\langle v_z^2 \rangle$ and $\langle v_\phi^2 \rangle$ which can be solved by means of one integration,
\begin{eqnarray}
n \langle v_z^2 \rangle (R,z) &=& \int_0^\infty n \frac{\partial \Phi}{\partial z} \diff z \label{eq:Jeans1}\\
n \langle v_\phi^2 \rangle (R,z) &=& R \frac{\partial}{\partial R} \left( \frac{n \langle v_z^2 \rangle}{1-\beta_z} \right) + \frac{n \langle v_z^2 \rangle}{1 - \beta_z} + R n \frac{\partial \Phi}{\partial R},\label{eq:Jeans2}
\end{eqnarray}
where $n(\vect{x}) \equiv \int f(\vect{x},\vect{v}) \Diff3 v$ is the number density of (stellar) tracers and $\Phi(\vect{x})$ the galaxy's gravitational potential generated by the mass density $\rho(\vect{x})$.

>>>>

%--------------------------------------------
A gravitational lens is a mass distribution whose gravitational potential $\Phi$ acts as a lens for light coming from a source positioned somewhere on a plane behind the lens. We summarize the basic formalism (see also \citealt{1992grle.book.....S,1996astro.ph..6001N,2006glsw.conf....1S,2006glsw.conf...91K,2010ARA&A..48...87T}). The angular diameter distance from the observer to the lens is $D_d$, to the source plane $D_s$, and the distance between the lens and source plane is $D_{ds}$. The deflection potential of the lens is its potential projected along the line of sight $z'$ and rescaled to
\begin{equation}
\psi(\vect{\theta}) \OLD{:=}\NEW{\equiv} \frac{D_{ds}}{D_d D_s} \frac{2}{c^2} \int \Phi(\vect{r}=D_d \vect{\theta},z') {\ \mathrm d} z', \label{eq:psidef}
\end{equation}
where $\vect{\theta}\equiv(x',y')$ is a 2-dimensional vector on the plane of the sky. The light from the source at $\vect{\beta} \equiv (x_s',y_s')$ is deflected according to the lens equation
\begin{equation}
\vect{\beta} = \vect{\theta}_i - \left.\vect{\nabla}_\theta \psi(\vect{\theta})\right|_{\vect{\theta}_i} \label{eq:lenseqpot}
\end{equation}
into an image $\vect{\theta}_i = (x_i',y_i')$. The gradient of the deflection potential, $\vect{\nabla}_\theta \psi(\vect{\theta})$, is equal to the angle by which the light is deflected multiplied by $D_{ds}/D_{s}$.

@@

The inverse magnification tensor
\begin{equation}
\mathscr{M}^{-1} \equiv \frac{\partial \vect{\beta}}{\partial \vect{\theta}} \overset{(\ref{eq:lenseqpot})}{=} \left(\delta_{ij} - \frac{\partial^2 \psi}{\partial \theta_i \partial \theta_j} \right)\label{eq:magnificationtensor}
\end{equation}
describes how the source position changes with image position\OLD{. It also describes}\NEW{,} the distortion of the image shape for an extended source and its magnification \OLD{$\mu$ due to lensing according to}\NEW{$\mu \equiv \text{image area}/\text{source area} = \det \mathscr{M}$}.
\OLD{$$\mu \equiv \frac{\text{image area}}{\text{source area}} = \det \mathscr{M}.$$}
Lines in the image plane for which the magnification becomes infinite, i.e., $\det \mathscr{M}^{-1} = 0$, are called critical curves. The corresponding lines in the source plane are called caustics. The position of the source with respect to the caustic detemines the number of images and their configuration and shape with respect to each other.

The Einstein mass $M_\text{ein}$ and Einstein radius $R_\text{ein}$ are defined via the relation
\begin{equation}
M_\text{ein} \equiv M(<R_\text{ein}) \overset{!}{=} \pi \Sigma_\text{crit} R_\text{ein}^2,
\end{equation}
where 
\begin{equation}
\Sigma_\text{crit} \equiv \frac{c^2}{4\pi G} \frac{D_s}{D_d D_{ds}}
\end{equation}
is the critical density and $M(<R_\text{ein})$ is the mass projected along the line-of-sight within $R'=R_\text{ein}$. $M_\text{ein}$ is similar to the projected mass within the critical curve $M_\text{crit}$.
%--------------------------------------------

@@

%--------------------------------------------
\Wilma{[Referee says: "The manuscript in its current form is also slightly longer than necessary at certain
points: [...] section
3.3.1 contains a lot of material that is implicit in Cappellari's JAM code, and
therefore does not need to be explained equation by equation here."]}
\OLD{JAM modelling assumes galaxies to be (i) collisionless (i.e., the collisionless Boltzmann equation for the distribution function $f(\vect{x},\vect{v},t)$ has to be satisfied, $\frac{\diff f(\vect{x},\vect{v},t)}{\diff t} = 0$), (ii) in a steady state ($\frac{\partial}{\partial t} = 0$), (iii) axisymmetric (best described in cylindrical coordinates $(R,z,\phi)$ and $\frac{\partial}{\partial \phi} = 0$). From this follow the axisymmetric Jeans equations as the vector-valued first moment of the Boltzmann equation, i.e.,
\begin{equation*}
\int \vect{v} \frac{\diff f}{\diff t} \Diff3 v = 0.
\end{equation*}}
\NEW{JAM modelling assumes galaxies to be (i) collisionless, (ii) in a steady state and (iii) axisymmetric. With these assumptions the axisymmetric Jeans equations follow from the vector-valued first moment of the collisionless Boltzmann equation (see Equations (4.221)-(4.222) in \Wilma{???}). The Jeans equations are functions of the second velocity moments $\langle v_i v_j\rangle$ (with $i,j\in(R,z,\phi)$), the stellar tracer number density \Wilma{Continue????????}} To be able to solve the Jeans equations, additional assumptions about the velocity ellipsoid tensor $\langle v_i v_j\rangle$ have to be made. We follow \citet{Cap08} and assume firstly that the galaxy's velocity ellipsoid is aligned with the cylindrical coordinate system, i.e., $\langle v_R v_z\rangle = \langle v_z v_\phi \rangle = \langle v_\phi v_R \rangle = 0$. Secondly, we assume a constant ratio between the radial and vertical second velocity moments, 
\begin{equation}
\beta_z \equiv 1 - \langle v_z^2 \rangle / \langle v_R^2\rangle, \label{eq:bz}
\end{equation}
where $\beta_z$ is the velocity anisotropy parameter. This reduces the Jeans equations to two equations for $\langle v_z^2 \rangle$ and $\langle v_\phi^2 \rangle$ which can be solved by means of one integration,
\begin{eqnarray}
n \langle v_z^2 \rangle (R,z) &=& \int_0^\infty n \frac{\partial \Phi}{\partial z} \diff z \label{eq:Jeans1}\\
n \langle v_\phi^2 \rangle (R,z) &=& R \frac{\partial}{\partial R} \left( \frac{n \langle v_z^2 \rangle}{1-\beta_z} \right) + \frac{n \langle v_z^2 \rangle}{1 - \beta_z} + R n \frac{\partial \Phi}{\partial R},\label{eq:Jeans2}
\end{eqnarray}
where $n(\vect{x}) \equiv \int f(\vect{x},\vect{v}) \Diff3 v$ is the number density of (stellar) tracers and $\Phi(\vect{x})$ the galaxy's gravitational potential generated by the mass density $\rho(\vect{x})$.
